Longitudinal Principal Manifold Estimation

### 2023-07-07 paper/main.tex
include application data description

To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice. In a larger comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability as measured by the Limits of Agreement were much higher for FIRST and FreeSurfer than for manual segmentation. \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains on which the segmentation models were trained [See original articles to confirm, see if there are any confirming studies].

@@

>>>>

To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of four years in the Alzheimer's Disease Neuroimaging Initiative (ADNI) dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice. In a larger comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability as measured by the Limits of Agreement were much higher for FIRST and FreeSurfer than for manual segmentation. \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains on which the segmentation models were trained [See original articles to confirm, see if there are any confirming studies].

@@

As discussed previously, meaningful between-image error in estimates of subcortical structures is introduced during the process of segmenting MRI images. This section demonstrates how the LPME algorithm may be used to mitigate the effects of such noise on further analysis using these estimates. To achieve this, we used MRI data collected through the ADNI study, a longitudinal observational study with the goal of identifying imaging biomarkers to assess the progression of Alzheimer's disease (AD).

While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, for the purposes of demonstration we used $T_1$-weighted structural MRI images from eight study participants in the cognitively normal group (\cite{jack2008adni}). Follow-up duration for these participants ranged between approximately 16 months to 48 months, with imaging scheduled to be conducted at six month intervals. Individuals from the cognitively normal group were selected because drastic changes in the volume and shape of subcortical structures over the course of the study follow-up duration would be unexpected. Under this assumption, we attribute volumetric changes in a given structure to error introduced by the study procedure or the image segmentation process. Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals. Images were processed using FSL and ANTs via the \texttt{fslr} and \texttt{ANTsR} packages in \texttt{R}. Specifically, ANTs was used for bias correcting the images, and the FIRST method in FSL was used for image segmentation.
